SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-04 design_arch.tex
Changes in Design. Minor edits in intro / method / results

As discussed in Section~\ref{sec:Background-Perceptron}, perceptrons
are useful, light-weight mechanisms to pull together disparate forms
of information and synthesize a decision from them.  In this case, we
will leverage a variety of different relative characteristics of a
given prefetch, such as speculation depth, page address and offset,
into a binary useful/non-useful prediction.  In the remainder of this
section, we discuss the the general structure of PPF for enhancing a
given underlying prefetcher.  We will show that PPF is a generalized
prefetch filtering mechanism which may adapted to any prefetcher with
minimal modifications. %Gino: Are these modifications minimal?

@@

The Perceptron Filter is organized as tables of perceptron weights,
with a separate table of weights for each feature.  A different number
of bits are used for each feature to index into its corresponding
table.
%Gino: repharse below?
Each feature's value is used to index into the corresponding table.
This indexing is done using a different number of bits for each
feature.  At most, 12 bits of a feature are used to index into the
weights table. Certain features require more resolution \textit{i.e.},
the full 12 bits of indexing.  On the other hand, some features
requiring lower resolution require as few as 7 bits of indexing.  The
variable indexing was determined by studying the features and
fine-tuned empirically to achieve a good accuracy vs hardware overhead
trade-off.  Exact details are mentioned under ``Area Overheads'' in
Section~\ref{Method-Overheads}. \pg{this whole ``bits'' discussion
  belongs in implementation not here IMHO}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15 and initialized
to zero at the beginning of a program's execution. Our proposed design
uses nine features, represented in nine perceptron weight
tables. \pg{the fact that its nine is again implementation details}

\pg{The word ChampSim should not appear in this part of the paper at
  all.  Remember you are describing how to build this in the general
  case...} \textbf{Inference}\newline In the ChampSim simulator, the
prefetching method is invoked on every L2 Cache demand access.  At
that point, the base prefetcher has the option to perform a prefetch
-- if it does, then it has a choice of how many cache lines to
prefetch.  The suggested prefetches can be either placed in the L2
cache or the last-level cache based on the prefetcher's internal
confidence mechanism.

@@

Once all the weights are retrieved, they are summed and compared to a
preset threshold, PERC\_ THRESHOLD\_LO.  Only the prefetch candidates
with perceptron sum higher than the threshold qualify for prefetching.
The prefetches that qualify through the perceptron stage are recorded
in the ``Prefetch Table''. The prefetch table is a 1024-entry, direct
mapped structure that contains all metadata required to recreate the
features.  %Gino: Recreate the features?
When the feedback of the current prefetched line is available at a
later stage, the stored data is used to train the perceptron.

%\pg{this discussion below has redundancy and doesn't flow well.
%  Rework it}

@@

%GAC: The section's previous contents are below, in case the above 
%rewritten work is not what we were looking for in terms of a rewording
%
%The perceptron sum can be considered the sum of the individual
%contribution of a feature.  The value of each contribution corresponds
%to a feature's confidence in the final decision on whether to prefetch
%a cache line. By summing the individual contributions, the final
%perceptron sum denotes the overall confidence for that prefetch
%suggestion.  By comparing the perceptron sum to two different
%thresholds, we divide the confidence scale of the prefetch suggestion
%into three bins. The first bin corresponds to the lowest confidence
%and leads to prefetch candidate being rejected.  The next bin
%corresponds to prefetches with a moderate confidence level.  This bin
%prefetches directeky to fill the bigger last level cache and prevents
%polluting the smaller L2 Cache.  The highest confidence prefetches are
%filled in the L2 cache.
%
%
%The sum of the perceptron weights are also used to decide whether to
%place the data in the L2 or the last level cache.  All prefetch
%candidates that are qualified until this stage are compared against
%PERC\_THRESHOLD\_HI to decide the fill level. The two thresholds:
%PERC\_THRESHOLD\_\_LO and PERC \_THRESHOLD\_HI are empirically set.

In addition to the prefetch table mentioned above, PPF also maintains
a ``Reject Table.''  The reject table is a 1024-entry deep
direct-mapped table.  If a prefetch suggestion is rejected by the
perceptron layer, it is logged into the reject table.  The filer is
used to train the perceptron to %Gino: Filer?
avoid false negatives \textit{i.e.}, cases where prediction was to
reject the prefetch but the prefetch would have been useful.

% djimenez: if it were previously stated then it's redundant here? do we need
% this statement?

@@

\textbf{Training}\newline In the prefetching environment, feedback for
a prefetch is received whenever there is an eviction or a demand
access from the L2 Cache.  This action triggers training of the base
prefetcher as well as the perceptron layer.  Training the perceptrons
involves accessing the metadata that was stored in Prefetch Table or
the Reject Table.  The cache lines address of the block being trained
is used to access a table, using 10-bits of the address for indexing
and another 6-bits to perform tag matching.  Once the state of the
program at the time of prefetching is available, it is
%Gino: By "state of the program" are you referring to the state of the features?
used to index into the perceptron weights table.

@@

If a cache block eviction led to training and the corresponding valid
entry is found in the prefetch table without the block being accessed,
then the prediction made by the perceptron was wrong.  The perceptron
should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to
reflect the misprediction. In either case, weights are saturated at
-16 or +15. %Gino: Describe how these limits were decided?

A secondary training mechanism also kicks in during demand fetches.
Before the demand access triggers the next set of prefetches, the
reject table is checked for a valid entry.  A hit means that the
corresponding cache line was initially suggested by the underlying
prefetch engine, but rejected by the perceptron filter.  Thus, the
perceptron should have been more confident about that particular
prefetch.  Once such a scenario is identified, the
state %Gino: state of execution == value of features?
of the execution at the time of prefetch is retrieved from the reject
table.  The retrieved data is used to index into the various weights
tables of the perceptron and the corresponding values are updated by
+1, saturating between and -16 and +15.  This update reflects
increased confidence for the prediction corresponding to that
prefetch.

@@

\item \textbf{Feature Selection:} In Section~\ref{Impl-Features}, we
  show the features used for PPF. Six out of the nine features we
  developed use information derived directly from program execution,
  irrespective of the baseline prefetcher. Beyond that, the feature
  set can be expanded to convey any useful information from the
  prefetcher to the perceptron filter.  The methodology explained in
  Section~\ref{Method-Features} talks about developing a minimal
  feature set for PPF. \pg{This should be more about how one should
    choose features than about what you did in the SPP implementation
    in particular.  Rewrite as a guide on picking features.}
%For example, in our implementation ``Confidence,'' ``Current Signature XOR Delta,''
%and ``Page Address XOR Confidence'' are such features.

>>>>

As discussed in Section~\ref{sec:Background-Perceptron}, perceptrons
are useful, light-weight mechanisms to pull together disparate forms
of information and synthesize a decision from them.  In this case, we
will leverage a variety of different relative characteristics of a
given prefetch, such as speculation depth, page address and offset,
into a binary useful/non-useful prediction.  In the remainder of this
section, we discuss the the general structure of PPF for enhancing a
given underlying prefetcher.  We will show that PPF is a generalized
prefetch filtering mechanism which may adapted to any prefetcher with
appropriate modifications.
% Gino: Are these modifications minimal?
% EB: That's a good question :P. Edited to 'appropriate'.

@@

%Gino: repharse below?
The Perceptron Filter is organized as tables of perceptron weights,
with a separate table of weights for each feature. For a design using 
`N' number of features, it's weights will be arranged in N different 
tables. Each feature's value is used to index into the corresponding table.
Since each table gets indexed independently, different number of bits can 
be used for indexing, thus giving each table a different depth.
The variable indexing can be determined by studying the features and
fine-tuned empirically to achieve a good accuracy vs hardware overhead
trade-off.  Exact details of our PPF implementation are mentioned 
under ``Area Overheads'' in Section~\ref{Method-Overheads}.

%\pg{this whole ``bits'' discussion belongs in implementation not here IMHO}
%[EB] Resolved

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15 and initialized
to zero at the beginning of a program's execution. It was empirically 
observed that 5-bits of indexing achieves a good trade-off between 
hardware overheads and accuracy. 
%\pg{the fact that its nine is again implementation details}
% [EB] Resolved

% \pg{The word ChampSim should not appear in this part of the paper at
%   all.  Remember you are describing how to build this in the general
%   case...} 
% [EB] Resolved
\textbf{Inference}\newline The prefetcher gets triggered on every 
L2 Cache demand access.  At
that point, the base prefetcher has the option to perform a prefetch
-- if it does, then it has a choice of how many cache lines to
prefetch.  The suggested prefetches can be either placed in the L2
cache or the last-level cache based on the prefetcher's internal
confidence mechanism.

@@

Once all the weights are retrieved, they are summed and compared to a
preset threshold, PERC\_ THRESHOLD\_LO.  Only the prefetch candidates
with perceptron sum higher than the threshold qualify for prefetching.
The prefetches that qualify through the perceptron stage are recorded
in the ``Prefetch Table''. The prefetch table is a 1024-entry, direct
mapped structure that contains all metadata required to re-index the 
perceptron entries for training.  
When the feedback of the current prefetched line is available at a
later stage, the stored data is used to train the perceptron.
%\pg{this discussion below has redundancy and doesn't flow well.
%  Rework it}

@@

%GAC: The section's previous contents are below, in case the above 
%rewritten work is not what we were looking for in terms of a rewording
%
%The perceptron sum can be considered the sum of the individual
%contribution of a feature.  The value of each contribution corresponds
%to a feature's confidence in the final decision on whether to prefetch
%a cache line. By summing the individual contributions, the final
%perceptron sum denotes the overall confidence for that prefetch
%suggestion.  By comparing the perceptron sum to two different
%thresholds, we divide the confidence scale of the prefetch suggestion
%into three bins. The first bin corresponds to the lowest confidence
%and leads to prefetch candidate being rejected.  The next bin
%corresponds to prefetches with a moderate confidence level.  This bin
%prefetches directly to fill the bigger last level cache and prevents
%polluting the smaller L2 Cache.  The highest confidence prefetches are
%filled in the L2 cache.
%
%
%The sum of the perceptron weights are also used to decide whether to
%place the data in the L2 or the last level cache.  All prefetch
%candidates that are qualified until this stage are compared against
%PERC\_THRESHOLD\_HI to decide the fill level. The two thresholds:
%PERC\_THRESHOLD\_\_LO and PERC \_THRESHOLD\_HI are empirically set.

In addition to the prefetch table mentioned above, PPF also maintains
a ``Reject Table.''  The reject table is a 1024-entry deep
direct-mapped table.  If a prefetch suggestion is rejected by the
perceptron layer, it is logged into the reject table.  The table is
used to train the perceptron to 
avoid false negatives \textit{i.e.}, cases where prediction was to
reject the prefetch but the prefetch would have been useful.

@@

\textbf{Training}\newline In the prefetching environment, feedback for
a prefetch is received whenever there is an eviction or a demand
access from the L2 Cache.  This action triggers training of the base
prefetcher as well as the perceptron layer.  Training the perceptrons
involves accessing the metadata that was stored in Prefetch Table or
the Reject Table.  The cache lines address of the block being trained
is used to access a table, using 10-bits of the address for indexing
and another 6-bits to perform tag matching.  Once the values of the
features at the time of prefetching is available, it is
used to index into the perceptron weights table.

@@

If a cache block eviction led to training and the corresponding valid
entry is found in the prefetch table without the block being accessed,
then the prediction made by the perceptron was wrong.  The perceptron
should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to
reflect the misprediction. In either case, weights are saturated at
-16 or +15.

A secondary training mechanism also kicks in during demand fetches.
Before the demand access triggers the next set of prefetches, the
reject table is checked for a valid entry.  A hit means that the
corresponding cache line was initially suggested by the underlying
prefetch engine, but rejected by the perceptron filter.  Thus, the
perceptron should have been more confident about that particular
prefetch.  Once such a scenario is identified, the
values of the features at the time of prefetch is retrieved from the reject
table.  The retrieved data is used to index into the various weights
tables of the perceptron and the corresponding values are updated by
+1, saturating between and -16 and +15.  This update reflects
increased confidence for the prediction corresponding to that
prefetch.

@@

\item \textbf{Feature Selection:}  Perceptrons essentially normalize
  contributions from different features to get a single sum representing the
  final confidence. So the perceptron learning can be as good as the set of 
  features chosen. Interestingly, this is what makes perceptron learning 
  scalable, as it can easily learn to incorporate newer information in 
  form of new features. Some of the features we developed use information 
  derived directly from program execution, agnostic to the baseline 
  prefetcher. Beyond that, the feature set can be expanded to convey 
  any useful information from the prefetcher to the perceptron filter.

  %[EB] This statement fits better in the next section 'Features used by Perceptron'
  %The methodology explained in
  %Section~\ref{Method-Features} talks about developing a minimal feature 
  %set for PPF.

%\pg{This should be more about how one should
%choose features than about what you did in the SPP implementation
%in particular.  Rewrite as a guide on picking features.}
% [EB]: Hopefully this is general enough and not too redundant

### 2018-12-04 design_impl.tex
Changes in Design. Minor edits in intro / method / results

\item \textbf{Enhanced Prefetch Table}\newline The prefetch table
  introduced in the original SPP keeps track of up to 1024 L2 cache
  prefetch suggestions.  For PPF, the prefetch table is modified to
  store the features used for perceptron indexing..
  This metadata is needed to retrieve the
  values of the features to train the perceptron at a later stage in
  time. The prefetch table mentioned above keeps track of the
  prefetches that actually happened in the L2 Cache.  We introduce the
  reject table to keep a track of the prefetches that were suggested
  by the base SPP engine but were rejected by PPF.  In case it is
  detected that a future demand fetch could have used this rejected
  prefetch, the perceptron weights are updated to reflect that
  misprediction.
%Gino: What's the etc.?
%EB: Resolved
\end{itemize}

\subsection{Features used by Perceptron}
\label{Impl-Features}
Here we discuss the various features that correlate the prefetching
decision with the program behavior. \pg{again you want to get at why
  these were chosen.  Also mention that you searched over a large
  space of possible features and these were the best.}

@@

% djimenez: this is a nice intuition and explanation.

@@

\item \textbf{Current Signature XOR Delta}: Recall from the discussion
  of SPP in Section \ref{Background-SPP} that the new signature is
  generated using the old signature and the current block delta.
%Gino: Added the sentence below, is this correct?
  The result of this feature is the next delta that is predicted to be
  accessed based on the delta predicted by SPP. While ``Current
  Signature XOR Delta'' is not the exact formula for generating the
  future signature, it gives an approximate idea of the path that the
  combination of these two values can lead to.

@@

>>>>

\item \textbf{Enhanced Prefetch Table}\newline The prefetch table
  introduced in the original SPP keeps track of up to 1024 L2 cache
  prefetch suggestions.  For PPF, the prefetch table is modified to
  store the features used for perceptron indexing..
  This metadata is needed to retrieve the
  values of the features to train the perceptron at a later stage in
  time. The prefetch table mentioned above keeps track of the
  prefetches that actually happened in the L2 Cache.  We introduce the
  reject table to keep a track of the prefetches that were suggested
  by the base SPP engine but were rejected by PPF.  In case it is
  detected that a future demand fetch could have used this rejected
  prefetch, the perceptron weights are updated to reflect that
  misprediction.
\end{itemize}

\subsection{Features used by Perceptron}
\label{Impl-Features}
Here we discuss the various features that correlate the prefetching
decision with the program behavior. All the features we used can be 
derived from the information available in the L2 Cache access stream.
Our feature selction involved searching over a large space of 
relevant perceptron features. Using the statistical methodology 
outlined in Section~\ref{Method-Features}, we pruned the feature 
set to a minimal yet relevant set of features
% \pg{again you want to get at why
% these were chosen.  Also mention that you searched over a large
% space of possible features and these were the best.}

% [EB] Tried addressing. Does it look reduntant when read after 'Feature Selection'
% bullet from the previous section?

@@

% djimenez: this is a nice intuition and explanation. 
% [EB] Thank you!

@@

\item \textbf{Current Signature XOR Delta}: Recall from the discussion
  of SPP in Section \ref{Background-SPP} that the new signature is
  generated using the old signature and the current block delta.
%Gino: Added the sentence below, is this correct?
%[EB]: It gives the next signature instead of next delta
  The result of this feature is the next signature that is predicted to be
  accessed based on the delta predicted by SPP. While ``Current
  Signature XOR Delta'' is not the exact formula for generating the
  future signature, it gives an approximate idea of the path that the
  combination of these two values can lead to.

@@

% [EB]: Think of more implementation specific details
% 7 to 12 bits of indexing
